repdata-004 : Peer Assesment 1 
========================================================


# Loading and preprocessing the data

* The data is read from the csv file.

```{r}
activityData <- read.csv(file="activity.csv");

head(activityData);
```


* Initially, $date is treated as a string factor.

```{r}

lapply(activityData, class)

```

* It's converted using basic functions.

```{r}

activityData$date <- as.Date(activityData$date)

class(activityData$date)
```

* We're going to need data with and without nas

```{r}

dataFiltered <- activityData[!is.na(activityData$steps),]
dataRaw      <- activityData

```


* We'll be 

# What is mean total number of steps taken per day?


```{r}
data <- dataFiltered 

hist(data$steps, col="green", main="Steps per day" , xlab="steps");
```

* Now for the concrete values:

```{r}
c( mean(data$steps) , median(data$steps))
```

# What is the average daily activity pattern?

Looking at it in terms of intervals:

* First we build a summary table:

```{r}

intervalData <- data.frame(interval = unique(activityData$interval));

for (i in 1:nrow(intervalData)) {
  subset <- data[data$interval == data[i,"interval"],];
  intervalData[i,"mean_all_days"] <- mean(subset$steps)
};

head(intervalData);

```

* Then we plot it:

```{r}
plot(intervalData$mean_all_days
     , type="l"
     , xlab="interval"
     , ylab="mean(steps per day"
     );
```

* Let's get a concrete number for the interval with the largest mean:

```{r}
intervalData[which.max(intervalData$mean_all_days),]
```

# Inputing missing values

Looks like all those NAs we filtered might cause problems.

* Let's get a count:

```{r}
nrow(dataRaw[is.na(dataRaw$steps),])
```

* We're going to normalize the results in a new copy of the raw data.

 
```{r}

estimated <- data.frame(dataRaw[0,]);
estimated <- rbind(estimated, dataRaw)

head(estimated)
```

* We'll fill the NA with the interval mean from before working with the new data.

```{r}

for (i in 1:nrow(intervalData) ) {
  interval <- intervalData[i,"interval"]
  meanVal <- intervalData[i,"mean_all_days"]
    
  rows <- which(is.na(estimated$steps))
  estimated[rows, "steps"] <- meanVal
}

data <- estimated

head(data)
```

* Now we'll plot the new data:

```{r}
hist(data$steps, col="green", main="Steps per day (estimated)" , xlab="steps");
```

* And recalulate the core values

```{r}
c( mean(data$steps) , median(data$steps))
```

* Including all that missing data had a significant effect on the mean (~1/6 the original value)

# Are there differences in activity patterns between weekdays and weekends?

* To categorize the date, we'll have to process the data into a new column.

```{r}

isWeekend <- function(x) {
  if( x == "domingo" || x == "sábado") "weekend" else "weekday";
}

data$day <- weekdays(data$date)
data$dayType <- lapply(data$day, isWeekend)

#make it a factor
data$dayType <- sapply(data$dayType, as.factor)

print(lapply(data, class))
```

* Tada: 

```{r}
head(data[data$dayType == "weekday",])
head(data[data$dayType == "weekend",])
```

* So now we need to graph them:

```{r}
library("lattice")

dayType <- data$dayType

xyplot( steps~interval | dayType, data=data, type="l", layout=c(1,2))
```
